Kindly Bent to Free Us

### 2019-07-09 kindly/lang.tex
notation of regions

A region $\region{x}{e}$ is annotated with its nesting $n$
and the variable $x$ that may be borrowed in $e$.

@@

\paragraph{Previously}

Let us first define our conventions and meta-syntactic variables.
Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
Variables, type variables and kind variables are noted
respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
are noted $\schm$ and $\kschm$.
In the rest of the text, we will write lists with a bar and an index, for
instance $\Multi[i]{\tau}$.
Kind constants are noted in bold capital letters, such as $\kaff$ for affine.

The syntax of \lang is presented in \cref{grammar} and follows
traditional ML languages with let bindings.
Borrows, noted $\borrow{e}$, are annotated with a specification indicating
if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
We also distinguish reborrows, which are noted $\reborrow{e}$.
Match are indexed with a specification which can be either $\&^b$
(which $b$ a borrow specification) or $id$, indicating if the match
should operate on borrows or not.
Regions, noted $\region{x}{e}$ are annotated with their nesting ($n$)
and the variable whose borrow they enclose ($x$).
%
$\tapp{t}{\Multi{\tau}}$ denotes a type application where
$\T t$ is an abstract type constructor.
Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
and $\borrowty{k}{\tau}$ respectively.
%
Kinds can be either a kind variable $\kvar$, or constants
(linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
by a nesting level represented by a member of $N \cup \{\infty\}$.
Type schemes and kind schemes are guarded by a set of constraints.
%
Finally, constraints are composed of a list of kind inequalities.

@@

Using these various rules, we can distribute the bindings every time an
expression has multiple sub-expressions such as pairs, function applications
and let bindings. For instance, in the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{x}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
we use the rule {\sc Susp} to detach a borrow on the left of the pair.
Naturally, we must here ensure that the borrow on the left of the pair does not
escape, which brings us to the typing rules for regions.

@@

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{x}{e}$ the region which allow borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

@@

We illustrate this on the expression $\region{x}{f(\borrow[\IBORROW]{c})}$
when $c$ is a linear channel.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).

>>>>

A region $\region{\Sone x\BORROW}{e}$ is annotated with its nesting $n$, the variable $x$ that may be borrowed in $e$, and the kind of borrow $\BORROW$.

@@

% \paragraph{Previously}

% Let us first define our conventions and meta-syntactic variables.
% Expressions, types and kinds are noted respectively $e$, $\tau$, and $k$.
% Variables, type variables and kind variables are noted
% respectively $x$, $\tvar$ and $\kvar$. Types and kind schemes
% are noted $\schm$ and $\kschm$.
% In the rest of the text, we will write lists with a bar and an index, for
% instance $\Multi[i]{\tau}$.
% Kind constants are noted in bold capital letters, such as $\kaff$ for affine.

% The syntax of \lang is presented in \cref{grammar} and follows
% traditional ML languages with let bindings.
% Borrows, noted $\borrow{e}$, are annotated with a specification indicating
% if a borrow is immutable ($\IBORROW$) or mutable ($\MBORROW$).
% We also distinguish reborrows, which are noted $\reborrow{e}$.
% Match are indexed with a specification which can be either $\&^b$
% (which $b$ a borrow specification) or $id$, indicating if the match
% should operate on borrows or not.
% Regions, noted $\region{\Sone x\BORROW}{e}$ are annotated with their nesting ($n$)
% and the variable whose borrow they enclose ($x$).
% %
% $\tapp{t}{\Multi{\tau}}$ denotes a type application where
% $\T t$ is an abstract type constructor.
% Arrow and borrow types are annotated with a kind, noted $\tau \tarr{k} \tau$
% and $\borrowty{k}{\tau}$ respectively.
% %
% Kinds can be either a kind variable $\kvar$, or constants
% (linear $\klin$, affine $\kaff$ or unrestricted $\kun$) indexed
% by a nesting level represented by a member of $N \cup \{\infty\}$.
% Type schemes and kind schemes are guarded by a set of constraints.
% %
% Finally, constraints are composed of a list of kind inequalities.

@@

Using these various rules, we can distribute the bindings every time an
expression has multiple sub-expressions such as pairs, function applications
and let bindings. For instance, in the
expression
$\letin{a}{\text{create}\ 10\ x}
{\introPair{\region[{}]{\Sone x\IBORROW}{\app{\text{length}}{\borrow[\IBORROW]{a}}}}{a}}$,
we use the rule {\sc Susp} to detach a borrow on the left of the pair.
Naturally, we must here ensure that the borrow on the left of the pair does not
escape, which brings us to the typing rules for regions.

@@

In order to provide good support for imperative programming, \affe allows
to borrow resources.
Such borrows are constrained to a given syntactic region. We note
$\region{\Sone x\BORROW}{e}$ the region which allow borrows of $x$ to be used in $e$
and where $n$ is the nesting of the region.
Borrows can not be used outside of a corresponding region.
These remarks lead to an
algorithm for automatic region annotation: start from each borrow, create
a region, and extend it as much as possible while respecting other borrows
and binders.
A complete specification of the algorithm is provided in \cref{regionannot}.
In the rest of this section, we will only consider terms
with fully annotated regions.

@@

We illustrate this on the expression $\region{\Sone x\IBORROW}{f(\borrow[\IBORROW]{c})}$
when $c$ is a linear channel.
The first step is to check that
$\svar[\IBORROW]{c}{\text{channel}}$ is in $\E$. We then define $\E'$ which
corresponds to $\E$ where the suspended binding is replaced by a borrow binding.
%
This is done by the borrow relation, noted $\lregion{C}{x}{\E}{\E'}$ and
whose main rules are shown in \cref{selectrules:region}.
In our case, we consider an immutable binding, so we replace the suspended
binding by $\bvar{\borrow[\IBORROW]{c}}{\borrowty[\IBORROW]{k}{\text{channel}}}$.
We must also check that $\entail{C}{(\kun_n\lk k\lk\kun_\infty)}$.
We can then typecheck the body of the region, $f(\borrow[\IBORROW]{c})$ in our example, and check that the borrow made available in the region does not escape.
This is done by the constraint $\Cleq{\tau}{\klin_{n-1}}$, which ensures
that the kind of $\tau$ does not contain anything whose kind is ``above $n$''
(for instance, $\borrowty[\IBORROW]{\kun_n}{\text{channel}}$ would have kind
$\kun_n$).
